---
title: "Paige 474 Final Project"
author: "Paige Cooper"
date: "2025-05-05"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Load Data
```{r}
# load and separate by semicolon
data <- read.csv("data.csv", sep=";")
head(data)


# remove enrolled
library("tidyverse")
dropout.data <- data %>% filter(Target %in% c("Dropout", "Graduate" ))
```

```{r}
# How many graduate, enrolled, dropout
summary(as.factor(data$Target))

#Distribution of Target

library(ggplot2)
library(RColorBrewer)
ggplot(data, aes(x = Target, fill = Target)) +
  geom_bar() +
  scale_fill_brewer(palette = "Paired") +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5) +
 labs(title = "Distribution of College Result (Target Variable)", x = "Result (Target)", y = "Count") +
  theme_minimal()
```


## Simple summary
```{r}
## Looks like a lot of integer values refering to different occupations etc. may need to use as.factor
summary(dropout.data[-37])
summary(as.factor(dropout.data$Target))
```
```{r}
# clean names

library(skimr)
clean_names <- c(
  "MaritalStatus",
  "ApplicationMode",
  "ApplicationOrder",
  "Course",
  "AttendanceTime", 
  "PrevQualification",
  "PrevQualificationGrade",
  "Nationality",
  "MotherQualification",
  "FatherQualification",
  "MotherOccupation",
  "FatherOccupation",
  "AdmissionGrade",
  "Displaced",
  "SpecialNeeds",
  "Debtor",
  "TuitionUpToDate",
  "Gender",
  "ScholarshipHolder",
  "AgeAtEnrollment",
  "InternationalStudent",
  "Units1stCredited",
  "Units1stEnrolled",
  "Units1stEvaluated",
  "Units1stApproved",
  "Units1stGrade",
  "Units1stWithoutEval",
  "Units2ndCredited",
  "Units2ndEnrolled",
  "Units2ndEvaluated",
  "Units2ndApproved",
  "Units2ndGrade",
  "Units2ndWithoutEval",
  "UnemploymentRate",
  "InflationRate",
  "GDP",
  "Target"
)

colnames(dropout.data) <- clean_names

```

```{r}
# establish factors
dropout.data$Target <- as.factor(dropout.data$Target)
dropout.data$MaritalStatus <- as.factor(dropout.data$MaritalStatus)
dropout.data$ApplicationMode <- as.factor(dropout.data$ApplicationMode)
dropout.data$Course <- as.factor(dropout.data$Course)
dropout.data$AttendanceTime <- as.factor(dropout.data$AttendanceTime)
dropout.data$PrevQualification <- as.factor(dropout.data$PrevQualification)
dropout.data$Nationality <- as.factor(dropout.data$Nationality)
dropout.data$MotherQualification <- as.factor(dropout.data$MotherQualification)
dropout.data$FatherQualification <- as.factor(dropout.data$FatherQualification)
dropout.data$MotherOccupation <- as.factor(dropout.data$MotherOccupation)
dropout.data$FatherOccupation <- as.factor(dropout.data$FatherOccupation)
dropout.data$Displaced <- as.factor(dropout.data$Displaced)
dropout.data$SpecialNeeds <- as.factor(dropout.data$SpecialNeeds)
dropout.data$Debtor <- as.factor(dropout.data$Debtor)
dropout.data$TuitionUpToDate <- as.factor(dropout.data$TuitionUpToDate)
dropout.data$Gender <- as.factor(dropout.data$Gender)
dropout.data$ScholarshipHolder <- as.factor(dropout.data$ScholarshipHolder)
dropout.data$InternationalStudent <- as.factor(dropout.data$InternationalStudent)
```


```{r}
# summary
skim(dropout.data)
```


## categorical plots
```{r}
# Gender
ggplot(dropout.data, aes(x = Gender, fill = Target)) +
  geom_bar(position = "dodge") +
  scale_fill_brewer(palette = "Set1") +
  geom_text(stat = "count", aes(label = ..count..),position = position_dodge(width = 0.9), vjust = -0.5) +
  labs(title = "Dropout vs Graduate by Gender",
       x = "Gender", y = "Count") +
  scale_x_discrete(label = c("Female", "Male")) + 
  theme_minimal()

# Attendance Time
ggplot(dropout.data, aes(x = AttendanceTime, fill = Target)) +
  geom_bar(position = "dodge") +
  scale_fill_brewer(palette = "Set1") +
  geom_text(stat = "count", aes(label = ..count..),position = position_dodge(width = 0.9), vjust = -0.5) +
  labs(title = "Dropout vs Graduate by Attendance Time",
       x = "Attendance Time", y = "Count") +
  scale_x_discrete(label = c("Evening", "Daytime")) + 
  theme_minimal()

# Fees Up to Date
ggplot(dropout.data, aes(x = TuitionUpToDate, fill = Target)) +
  geom_bar(position = "dodge") +
  scale_fill_brewer(palette = "Set1") +
  geom_text(stat = "count", aes(label = ..count..),position = position_dodge(width = 0.9), vjust = -0.5) +
  labs(title = "Dropout vs Graduate by Tuition Fees Up to Date",
       x = "Tuition Fees Up to Date", y = "Count") +
  scale_x_discrete(label = c("Not Up to Date", "Up to Date")) + 
  theme_minimal()


# Scholarship Holder
ggplot(dropout.data, aes(x = ScholarshipHolder, fill = Target)) +
  geom_bar(position = "dodge") +
  scale_fill_brewer(palette = "Set1") +
  geom_text(stat = "count", aes(label = ..count..), position = position_dodge(width = 0.9), vjust = -0.5) +
  labs(title = "Dropout vs Graduate by Scholarship Holder",
       x = "Scholarship Holder", y = "Count") +
  scale_x_discrete(label = c("No Scholarship", "Scholarship")) + 
  theme_minimal()
```


## numeric plots
```{r}
# Age
ggplot(dropout.data, aes(x = Target, y = AgeAtEnrollment, fill = Target)) +
  geom_boxplot() +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Age at Enrollment by Target",
       y = "Age", x = "Target") +
  theme_minimal()

# Age
ggplot(dropout.data, aes(x = Target, y = AdmissionGrade, fill = Target)) +
  geom_boxplot() +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Admission Grade by Target",
       y = "Admission Grade", x = "Target") +
  theme_minimal()
```


## LDA 
Can correctly predict Grad versus dropout by 87.3%.
```{r}
train <- 1:2904
test <- 2905:3630


library(MASS)

# LDA based on only numerical predictors
lda.fit <-lda(Target ~ ApplicationOrder + PrevQualificationGrade + AdmissionGrade + AgeAtEnrollment + Units1stCredited + Units1stEnrolled + Units1stEvaluated + Units1stApproved + Units1stGrade + Units1stWithoutEval + 	Units2ndCredited + Units2ndEnrolled + Units2ndEvaluated + Units2ndApproved + Units2ndGrade + Units2ndWithoutEval + UnemploymentRate + InflationRate + GDP, data = dropout.data, subset = train)

# confusion matrix
targett <- dropout.data$Target[test]
lda.pred <- predict(lda.fit, dropout.data[test, ])
table(lda.pred$class, targett)

# correct percentage
634/726

```


## LDA with dummy variables
Can correctly predict Grad versus dropout by 89.8% when you include dummy variables for 0-1 predictors.
```{r}
train <- 1:2904
test <- 2905:3630


# LDA 
lda.fit <-lda(Target ~ ApplicationOrder + PrevQualificationGrade + AdmissionGrade + AgeAtEnrollment + Units1stCredited + Units1stEnrolled + Units1stEvaluated + Units1stApproved + Units1stGrade + Units1stWithoutEval + 	Units2ndCredited + Units2ndEnrolled + Units2ndEvaluated + Units2ndApproved + Units2ndGrade + Units2ndWithoutEval + UnemploymentRate + InflationRate + GDP + AttendanceTime + Displaced + SpecialNeeds + Gender + Debtor + TuitionUpToDate + ScholarshipHolder + InternationalStudent, data = dropout.data, subset = train)

# confusion matrix
targett <- dropout.data$Target[test]
lda.pred <- predict(lda.fit, dropout.data[test, ])
table(lda.pred$class, targett)

# correct percentage
(238+414)/726

lda.fit

```



## QDA
Performs slightly worse with 86.9% accuracy.
```{r}
# QDA based on only numerical predictors
qda.fit <-qda(Target ~ ApplicationOrder + PrevQualificationGrade + AdmissionGrade + AgeAtEnrollment + Units1stCredited + Units1stEnrolled + Units1stEvaluated + Units1stApproved + Units1stGrade + Units1stWithoutEval + 	Units2ndCredited + Units2ndEnrolled + Units2ndEvaluated + Units2ndApproved + Units2ndGrade + Units2ndWithoutEval + UnemploymentRate + InflationRate + GDP, data = dropout.data, subset = train)

# confusion matrix
targett <- dropout.data$Target[test]
qda.pred <- predict(qda.fit, dropout.data[test, ])
table(qda.pred$class, targett)

# correct percentage
631/ 726
```


## QDA w dummy variables
Performs slightly worse than LDA with dummy variables with 87.6% accuracy. Better than both qda and lda with no dummy variables.
```{r}
# QDA based on only numerical predictors
qda.fit <-qda(Target ~ ApplicationOrder + PrevQualificationGrade + AdmissionGrade + AgeAtEnrollment + Units1stCredited + Units1stEnrolled + Units1stEvaluated + Units1stApproved + Units1stGrade + Units1stWithoutEval + 	Units2ndCredited + Units2ndEnrolled + Units2ndEvaluated + Units2ndApproved + Units2ndGrade + Units2ndWithoutEval + UnemploymentRate + InflationRate + GDP + AttendanceTime + Displaced + SpecialNeeds + Gender + Debtor + TuitionUpToDate + ScholarshipHolder + InternationalStudent, data = dropout.data, subset = train)

# confusion matrix
targett <- dropout.data$Target[test]
qda.pred <- predict(qda.fit, dropout.data[test, ])
table(qda.pred$class, targett)

# correct percentage
(239 + 397)/ 726
```



## Testing Logistic - Look to see if we want to change logistic model. Now 90.9% accurate
```{r}
data <- read.csv("data.csv", sep=";")
head(data)

# remove enrolled
library("tidyverse")
dropout.data <- data %>% filter(Target %in% c("Dropout", "Graduate" ))

train <- 1:2904
test <- 2905:3630


dropout.data$Target <- factor(dropout.data$Target, levels = c("Graduate", "Dropout"))

logit.fit <- glm(Target ~ Application.order + Previous.qualification..grade. + Admission.grade + Age.at.enrollment +
                   Curricular.units.1st.sem..credited. + Curricular.units.1st.sem..enrolled. + Curricular.units.1st.sem..evaluations. +
                   Curricular.units.1st.sem..approved. + Curricular.units.1st.sem..grade. + Curricular.units.1st.sem..without.evaluations. +
                   Curricular.units.2nd.sem..credited. + Curricular.units.2nd.sem..enrolled. + Curricular.units.2nd.sem..evaluations. +
                   Curricular.units.2nd.sem..approved. + Curricular.units.2nd.sem..grade. + Curricular.units.2nd.sem..without.evaluations. +
                   Unemployment.rate + Inflation.rate + GDP + as.factor(Daytime.evening.attendance.) + as.factor(Displaced) + as.factor(Educational.special.needs) + as.factor(Gender) + as.factor(Debtor) + as.factor(Tuition.fees.up.to.date) + as.factor(Scholarship.holder) + as.factor(International),
                 data = dropout.data,
                 family = binomial,
                 subset = train)

#Predicting dropout probabilities on the test set, and this gives predicted probabilities between 0 and 1. We're also converting probabilities to class predictions. Threshold = 0.5: If prob > 0.5 then Dropout; else Graduate
logit.probs <- predict(logit.fit, newdata = dropout.data[test, ], type = "response")
logit.pred <- ifelse(logit.probs > 0.5, "Dropout", "Graduate")
logit.pred <- factor(logit.pred, levels = c("Graduate", "Dropout"))

# here I've added a confusion matrix to compare predictions vs actual values
table(Predicted = logit.pred, Actual = dropout.data$Target[test])
mean(logit.pred == dropout.data$Target[test])

summary(logit.fit)         # shows estimates and p-values

exp(coef(logit.fit)) 
```

